---
title: "Untitled"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
counts <- read.csv("airway_scaledcounts.csv", stringsAsFactors = FALSE)
metadata <-  read.csv("airway_metadata.csv", stringsAsFactors = FALSE)
```

```{r}
head(counts)
```

```{r}
head(metadata)
```

```{r}
colnames(counts)
```

```{r}
metadata$id
```


```{r}
all(colnames(counts)[-1] == metadata$id)
```

First Task: Find which columns in **counts** correspond to the control conditions (i.e. no drug!). Then we can calculate the mean count value for each gene.

```{r}
control.inds <- metadata$dex == "control"
control.ids <- metadata$id[control.inds]
head(counts[,control.ids])
```

Calculate the mean counts value for all genes
```{r}
control.mean <- rowSums(counts[,control.ids])/ncol(counts[,control.ids])
```

Now do the same thing for the drug treated samples (i.e. **treated** columns)

```{r}
treated.inds <- metadata$dex == "treated"
treated.ids <- metadata$id[treated.inds]
head(counts[,treated.ids])
treated.mean <- rowSums(counts[,treated.ids])/ncol(counts[,treated.ids])
 
names(treated.mean) <- counts$ensgene
```

Combine the mean count values for control and treated into a new data.frame called meancounts!
```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

```{r}
colSums(meancounts)
```

Now we can plot the drug treated vs non-drug control count values
```{r}
plot(meancounts$control.mean, meancounts$treated.mean)
```


```{r}
plot(meancounts$control.mean, meancounts$treated.mean, log="xy")
```

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

```{r}
toy <- c(1,2,0,4,0)
which(toy==0)
```

```{r}
toy <- meancounts[1:6,]
toy[,1:2]==0
unique(which(toy[,1:2]==0, arr.ind = TRUE)[,1])
```


```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

```{r}
sum(up.ind)
```

```{r}
sum(down.ind)
```

```{r}
head(mycounts[up.ind,])
```


There are `sum(up.ind)` up regulated genes and `sum(down.ind)` down regulated genes.

```{r}
anno <- read.csv("annotables_grch38.csv")
head(anno)
```

```{r}
mycounts.annotated <- merge(x=mycounts, y=anno, by.x="row.names", by.y="ensgene")
head(mycounts.annotated)
```

##DESeq2 analysis

Now we do this analysis properly with p-values!

```{r}
library(DESeq2)
```

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex, 
                              tidy=TRUE)
dds
```

```{r}
dds <- DESeq(dds)
res <- results(dds)
res
```

```{r}
summary(res)
```

```{r}
res01 <- results(dds, alpha = 0.01)
summary(res01)
```

```{r}
plot(res01$log2FoldChange, -log(res01$padj))
```

```{r}
mycols <- rep("gray", nrow(res01))
mycols[abs(res01$log2FoldChange)>2] <- "red"
inds <- (res01$padj<0.01)&(abs(res01$log2FoldChange)>2)
mycols[inds] <- "blue"
#Volcano plot with custom colors
plot(res01$log2FoldChange, -log(res01$padj), col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)")
abline(v=c(-2, 2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```














